---
title: "Basic Classifier for Storm Event Type"
output: html_document
---

# Jason Ryan May

**Research Question:**

* Given where a storm occurs, when it occurs, what the local climate and demographics look like, and the storm's physical characteristics; Can we predict what type of severe weather event it is?


# Data Cleaning and Preprocessing

```{r, include=FALSE}
library(tidyverse)
library(tidymodels)
```

```{r}
# Load Raw Data
raw_df <- read_csv("mergedData_2023.csv", show_col_types = FALSE)
raw_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```

* It should be noted there is an imbalance of observations in each class. Additionally, there are only 257 lighting events, which is a very small number of observations to train a model on.


```{r}
# DAMAGE_PROPERTY (71.5% zeros), STORM_AREA_SQMILES (79.9% zeros), and
# DURATION_MINUTES (57.8% zeros) are heavily zero-inflated. Adding binary
# flags lets the model split on "zero vs nonzero" independently from
# the magnitude.
 
# create new binary columns for the zero-inflated variables
new_df <- raw_df |>
  mutate(
    DAMAGE_PROPERTY_NONZERO    = as.integer(DAMAGE_PROPERTY > 0),
    STORM_AREA_SQMILES_NONZERO = as.integer(STORM_AREA_SQMILES > 0),
    DURATION_MINUTES_NONZERO   = as.integer(DURATION_MINUTES > 0)
  )

# drop the original columns since the magnitude is not meaningful when zero-inflated
new_df <- new_df |>
  select(-DAMAGE_PROPERTY, -STORM_AREA_SQMILES, -DURATION_MINUTES)

# drop state, month_name, FIPS and cz_name since they are redundant with the lat and lon and month number
new_df <- new_df |>
  select(-STATE, -MONTH_NAME, -CZ_NAME, -FIPS)


# hot swap all categorical variables to numeric factors
new_df <- new_df |>
  mutate(
    EVENT_TYPE = factor(EVENT_TYPE),
    COASTAL_TYPE_SHORELINE = as.numeric(as.factor(COASTAL_TYPE_SHORELINE)),
    COASTAL_TYPE_WATERSHED = as.numeric(as.factor(COASTAL_TYPE_WATERSHED)),
    MODAL_YEAR_BUILT_BIN   = as.numeric(as.factor(MODAL_YEAR_BUILT_BIN))
    )

glimpse(new_df)
```

```{r}
# Lightning has too few observations (~256) and is nearly indistinguishable
# from Thunderstorm Wind using these features. Merging improves model
# performance on all other classes without meaningful information loss.
 
new_df <- new_df |>
  mutate(EVENT_TYPE = if_else(EVENT_TYPE == "Lightning",
                              "Thunderstorm Wind",
                              EVENT_TYPE))
 
cat("\nClass distribution after merging Lightning into Thunderstorm Wind:\n")
new_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```


```{r}
# Stratified Train / Test Split (80/20)

set.seed(777)
data_split <- initial_split(new_df, prop = 0.8, strata = EVENT_TYPE)
train_data <- training(data_split)
test_data  <- testing(data_split)
 
cat("\nTrain size:", nrow(train_data), "| Test size:", nrow(test_data), "\n")
 
cat("\nTrain class distribution:\n")
train_data |> count(EVENT_TYPE, sort = TRUE) |> print()
```

# Model training -- Random Forest Training & Tuning

```{r, include = FALSE}
library(ranger)
```

```{r}
# set the recipe
rf_recipe <- recipe(EVENT_TYPE ~ ., data = train_data) |>
  step_zv(all_predictors())
```

```{r}

# setting model specifications (the recipe proportions)
rf_spec <- rand_forest(
  trees = 300,
  mtry  = tune(),
  min_n = tune()
) |>
  set_engine("ranger", max.depth = 15, importance = "impurity") |>
  set_mode("classification")
 
```

```{r}

# setting the workflow (how to make the recipe)
rf_workflow <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)

```

```{r}
# Cross-Validation Folds

set.seed(777)

# Stratified 5-fold cross-validation ensures each fold has a representative distribution of the EVENT_TYPE classes, which is crucial given the class imbalance.

cv_folds <- vfold_cv(train_data, v = 5, strata = EVENT_TYPE)

```

```{r, cache = TRUE}
# tuning

n_features <- ncol(train_data) - 1 # subtract 1 for the target variable
sqrt_features <- floor(sqrt(n_features))

# Grid search .
rf_grid <- grid_regular(
  mtry(range = c(sqrt_features, n_features)),
  min_n(range = c(5, 15)),
  levels = 4
)

# Macro f1 is the primary metric for evaluation due to the class imbalance.
tune_results <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(f_meas),
  control   = control_grid(verbose = FALSE)
)
```

```{r}
cat("\nTop 5 configurations by macro-F1:\n")
top_5_params <- tune_results |>
  show_best(metric = "f_meas", n = 5)

print(top_5_params)

best_params <- select_best(tune_results, metric = "f_meas")
cat("\nBest hyperparameters:\n")
print(best_params)

# use 7 and 6 instead
# manual_best_params <- tibble(
#   mtry = 11,
#   min_n = 6,
# )
```

```{r}
# fit worlflow with the best hyperparameters on the full training set
final_workflow <- rf_workflow |>
  finalize_workflow(best_params)
```

```{r, cache=TRUE}
cat("\nFitting final model on full training set...\n")
final_model <- final_workflow |>
  fit(data = train_data)
```



# Evaluation
```{r, include = FALSE}
library(vip)
```

```{r}
# add the predictions to the train and test for evaluation and reporting 

train_preds <- train_data |>
  bind_cols(predict(final_model, train_data))

# mutate event type to factor data type
train_preds <- train_preds |>
  mutate(EVENT_TYPE = factor(EVENT_TYPE))

test_preds <- test_data |>
  bind_cols(predict(final_model, test_data))

# mutate event type to factor data type
test_preds <- test_preds |>
  mutate(EVENT_TYPE = factor(EVENT_TYPE))
```


```{r}
# Accuracy check

# training set
train_accuracy <- mean(train_preds$.pred_class == train_preds$EVENT_TYPE)

bal_acc_train <- train_preds |>
  bal_accuracy(truth = EVENT_TYPE, estimate = .pred_class) |>
  pull(.estimate)

cat("Training Accuracy:  ", round(train_accuracy, 3), "\n")
cat("Balanced Training Accuracy: ", round(bal_acc_train, 3), "\n")


# test set
test_accuracy <- mean(test_preds$.pred_class == test_preds$EVENT_TYPE)

bal_acc_test <- test_preds |>
  bal_accuracy(truth = EVENT_TYPE, estimate = .pred_class) |>
  pull(.estimate)

cat("Test Accuracy:  ", round(test_accuracy, 3), "\n")
cat("Balanced Test Accuracy: ", round(bal_acc_test, 3), "\n")
```

```{r}
conf_mat_result <- test_preds |>
  conf_mat(truth = EVENT_TYPE, estimate = .pred_class)
print(conf_mat_result)
```


```{r}
# Per Class Prediction metrics
 
# Recall: the ability to find all actual positive cases (TP / TP + FN)
recall_tbl <- test_preds |>
  group_by(EVENT_TYPE) |>
  summarise(
    n       = n(),
    correct = sum(EVENT_TYPE == .pred_class),
    recall  = correct / n,
    .groups = "drop"
  )
 
# Precision: the accuracy of positive predictions (TP / TP + FP)
precision_tbl <- test_preds |>
  group_by(.pred_class) |>
  summarise(
    predicted_n = n(),
    true_pos    = sum(EVENT_TYPE == .pred_class),
    precision   = true_pos / predicted_n,
    .groups     = "drop"
  ) |>
  rename(EVENT_TYPE = .pred_class)
 
# Combine into a single report
class_report <- recall_tbl |>
  left_join(precision_tbl |> select(EVENT_TYPE, precision), by = "EVENT_TYPE") |>
  mutate(f1 = 2 * (precision * recall) / (precision + recall)) |>
  select(EVENT_TYPE, n, precision, recall, f1) |>
  arrange(desc(f1))
 
cat("\nPer-Class Metrics:\n")
print(class_report)
 
macro_f1    <- mean(class_report$f1, na.rm = TRUE)
weighted_f1 <- weighted.mean(class_report$f1, class_report$n, na.rm = TRUE)

cat("\nMacro-average F1:   ", round(macro_f1, 3), "\n")
cat("Weighted-average F1:", round(weighted_f1, 3), "\n")
```

```{r}
# Feature Importance
 
final_fitted <- final_model |> extract_fit_parsnip()
 
importance_tbl <- final_fitted |>
  vip::vi() |>
  arrange(desc(Importance))
 
cat("Feature Importances (Gini impurity):\n")
print(importance_tbl, n = 20)
```

```{r}
# Feature Importance Plot
 
importance_tbl |>
  mutate(Variable = fct_reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col(fill = "steelblue") +
  labs(
    title    = "Random Forest — Feature Importance",
    subtitle = "Storm Event Type Classification (Gini Impurity)",
    x        = "Importance",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)

```

```{r}
# Confusion Matrix Heatmap
 
cm_tbl <- conf_mat_result$table |>
  as_tibble() |>
  rename(Predicted = Prediction, Actual = Truth, Count = n)
 
cm_tbl |>
  ggplot(aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Count), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix — Random Forest",
    x     = "Predicted Class",
    y     = "Actual Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

```

# Extra Graphics

```{r, include = FALSE}
library(rpart)
library(rpart.plot)
library(patchwork)
library(maps)
```

```{r}
tree_fit <- rpart(
  EVENT_TYPE ~ .,
  data    = train_data,
  method  = "class",
  control = rpart.control(
    maxdepth = 4,
    cp       = 0.005,
    minsplit = 100
  )
)

rpart.plot(
  tree_fit,
  type          = 3,
  extra         = 102,
  under         = TRUE,
  fallen.leaves = TRUE,
  roundint      = FALSE,
  cex           = 0.7,
  box.palette   = "BuGn",
  main          = "Surrogate Decision Tree — Storm Event Type Classification",
  sub           = "Simplified single tree showing the splitting logic used by the Random Forest",
  legend.x = NA
)
```

```{r, cache=TRUE}
# Manually computed PDPs for the top 4 features, showing how each shifts the predicted class probabilities.
 
prepped_recipe <- prep(rf_recipe, training = train_data)
train_baked <- bake(prepped_recipe, new_data = train_data) |> select(-EVENT_TYPE)
 
fitted_ranger <- extract_fit_parsnip(final_model)$fit
 
top_features <- c("STORM_AREA_SQMILES_NONZERO", "BEGIN_LON", "TEMPERATURE_F", "BEGIN_LAT")
 
# Get class names — try multiple sources in order of reliability
class_levels <- fitted_ranger$forest$levels
if (is.null(class_levels) || length(class_levels) == 0) {
  class_levels <- levels(train_data$EVENT_TYPE)
}
if (is.null(class_levels) || length(class_levels) == 0) {
  class_levels <- sort(unique(as.character(train_data$EVENT_TYPE)))
}
 
compute_pdp <- function(model, data, feature, grid_size = 40) {
  feat_vals <- data[[feature]]
  n_na <- sum(is.na(feat_vals))
  if (n_na > 0) cat("    Note:", n_na, "NAs in", feature, "(ignored)\n")
 
  grid_vals <- seq(
    quantile(feat_vals, 0.02, na.rm = TRUE),
    quantile(feat_vals, 0.98, na.rm = TRUE),
    length.out = grid_size
  )
 
  pdp_results <- map_dfr(grid_vals, function(val) {
    modified <- data
    modified[[feature]] <- val
    preds <- predict(model, data = as.data.frame(modified), type = "response")
    avg_probs <- colMeans(preds$predictions, na.rm = TRUE)
    # Use prediction column names if class_levels failed
    pred_names <- if (length(class_levels) > 0) class_levels else colnames(preds$predictions)
    tibble(value = val) |>
      bind_cols(as_tibble_row(setNames(as.list(avg_probs), pred_names)))
  })
 
  pdp_results |>
    pivot_longer(-value, names_to = "EVENT_TYPE", values_to = "Probability") |>
    mutate(feature = feature)
}
 
pdp_data <- map_dfr(top_features, function(feat) {
  cat("  Computing PDP for", feat, "...\n")
  compute_pdp(fitted_ranger, train_baked, feat)
})
 
pdp_data |>
  ggplot(aes(x = value, y = Probability, color = EVENT_TYPE)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ feature, scales = "free_x", ncol = 2) +
  scale_color_brewer(palette = "Set2") +
  labs(
    title    = "Partial Dependence Plots — Top 4 Features",
    subtitle = "How each feature shifts predicted probability by event type",
    x        = "Feature Value",
    y        = "Predicted Probability",
    color    = "Event Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )
```

```{r}

# Plots each storm event as a point on a CONUS map, colored by event type.
# Reveals spatial clustering that lat/lon captures for the model.
 
us_map <- map_data("state")
 
# Combine train and test for a complete picture
all_data <- bind_rows(train_data, test_data)
 
# Sample down for readability
set.seed(777)
plot_sample <- all_data |>
  group_by(EVENT_TYPE) |>
  slice_sample(n = 1500) |>
  ungroup()
 
ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey95", color = "grey70", linewidth = 0.3
  ) +
  geom_point(
    data = plot_sample,
    aes(x = BEGIN_LON, y = BEGIN_LAT, color = EVENT_TYPE),
    alpha = 0.4, size = 0.6
  ) +
  facet_wrap(~ EVENT_TYPE, ncol = 3) +
  scale_color_brewer(palette = "Set2") +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(
    title    = "Distribution of Storm Events (2023)",
    subtitle = "Each point is a storm event, faceted by type",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank()
  )
```

```{r}
# Creates a grid of lat/lon points across CONUS, predicts class
# probabilities at each point (holding other features at their medians),
# and plots the most-likely class and its confidence as a filled map.
 
# Build a grid of geographic points
lat_grid <- seq(25, 49, length.out = 60)
lon_grid <- seq(-124, -67, length.out = 80)
geo_grid <- crossing(BEGIN_LAT = lat_grid, BEGIN_LON = lon_grid)
 
# Fill all other features with their training set medians
median_values <- train_baked |>
  select(-BEGIN_LAT, -BEGIN_LON) |>
  summarise(across(everything(), median))
 
grid_data <- geo_grid |>
  bind_cols(median_values |> slice(rep(1, nrow(geo_grid))))
 
# Predict class probabilities at each grid point
grid_preds <- predict(
  fitted_ranger,
  data = as.data.frame(grid_data),
  type = "response"
)
 
# Extract the most likely class and its probability
pred_probs <- as_tibble(grid_preds$predictions)
colnames(pred_probs) <- class_levels
 
grid_results <- geo_grid |>
  bind_cols(pred_probs) |>
  mutate(
    predicted_class = class_levels[max.col(pred_probs)],
    confidence      = apply(pred_probs, 1, max)
  )
 
# Plot: tile colored by predicted class, alpha by confidence
grid_results |>
  ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = NA, color = "grey40", linewidth = 0.3
  ) +
  geom_tile(
    aes(x = BEGIN_LON, y = BEGIN_LAT,
        fill = predicted_class, alpha = confidence),
    width = 0.8, height = 0.45
  ) +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = NA, color = "grey40", linewidth = 0.3
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_alpha_continuous(range = c(0.3, 1.0), guide = "none") +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(
    title    = "Predicted Storm Type Across CONUS",
    subtitle = "Based on lat/lon only (all other features held at median). Opacity = model confidence.",
    fill     = "Predicted\nEvent Type",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank()
  )
```

```{r}
 
# Generate predictions on the test set
test_with_preds <- test_data |>
  bind_cols(predict(final_model, test_data)) |>
  mutate(correct = EVENT_TYPE == .pred_class)

# sample down for readability
set.seed(777)
test_with_preds <- test_with_preds |>
  group_by(EVENT_TYPE) |>
  slice_sample(n = 1500) |>
  ungroup()

 
# Faceted map: predicted class
ggplot() +
  geom_polygon(
    data = us_map,
    aes(x = long, y = lat, group = group),
    fill = "grey95", color = "grey70", linewidth = 0.3
  ) +
  geom_point(
    data = test_with_preds,
    aes(x = BEGIN_LON, y = BEGIN_LAT, color = .pred_class),
    alpha = 0.4, size = 0.6
  ) +
  facet_wrap(~ .pred_class, ncol = 3) +
  scale_color_brewer(palette = "Set2") +
  coord_quickmap(xlim = c(-125, -66), ylim = c(24, 50)) +
  labs(
    title    = "Predicted Storm Types — Test Set",
    subtitle = "Each point is a test event, colored and faceted by model prediction",
    x        = "Longitude",
    y        = "Latitude"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    legend.position  = "none",
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank()
  )

```

