---
title: "Basic Classifier for Storm Event Type"
output: html_document
---

# Jason Ryan May

**Research Question:**

* Given where a storm occurs, when it occurs, what the local climate and demographics look like, and the storm's physical characteristics; Can we predict what type of severe weather event it is?


# Data Cleaning and Preprocessing

```{r, include=FALSE}
library(tidyverse)
library(tidymodels)
```

```{r}
# Load Raw Data
raw_df <- read_csv("mergedData_2023.csv", show_col_types = FALSE)
raw_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```

* It should be noted there is an imbalance of observations in each class. Additionally, there are only 257 lighting events, which is a very small number of observations to train a model on.


```{r}
# DAMAGE_PROPERTY (71.5% zeros), STORM_AREA_SQMILES (79.9% zeros), and
# DURATION_MINUTES (57.8% zeros) are heavily zero-inflated. Adding binary
# flags lets the model split on "zero vs nonzero" independently from
# the magnitude.
 
# create new binary columns for the zero-inflated variables
new_df <- raw_df |>
  mutate(
    DAMAGE_PROPERTY_NONZERO    = as.integer(DAMAGE_PROPERTY > 0),
    STORM_AREA_SQMILES_NONZERO = as.integer(STORM_AREA_SQMILES > 0),
    DURATION_MINUTES_NONZERO   = as.integer(DURATION_MINUTES > 0)
  )

# drop the original columns since the magnitude is not meaningful when zero-inflated
new_df <- new_df |>
  select(-DAMAGE_PROPERTY, -STORM_AREA_SQMILES, -DURATION_MINUTES)

# drop state, month_name, FIPS and cz_name since they are redundant with the lat and lon and month number
new_df <- new_df |>
  select(-STATE, -MONTH_NAME, -CZ_NAME, -FIPS)


# hot swap all categorical variables to numeric factors
new_df <- new_df |>
  mutate(
    EVENT_TYPE = factor(EVENT_TYPE),
    COASTAL_TYPE_SHORELINE = as.numeric(as.factor(COASTAL_TYPE_SHORELINE)),
    COASTAL_TYPE_WATERSHED = as.numeric(as.factor(COASTAL_TYPE_WATERSHED)),
    MODAL_YEAR_BUILT_BIN   = as.numeric(as.factor(MODAL_YEAR_BUILT_BIN))
    )

glimpse(new_df)
```

```{r}
# Lightning has too few observations (~256) and is nearly indistinguishable
# from Thunderstorm Wind using these features. Merging improves model
# performance on all other classes without meaningful information loss.
 
new_df <- new_df |>
  mutate(EVENT_TYPE = if_else(EVENT_TYPE == "Lightning",
                              "Thunderstorm Wind",
                              EVENT_TYPE))
 
cat("\nClass distribution after merging Lightning into Thunderstorm Wind:\n")
new_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```


```{r}
# Stratified Train / Test Split (80/20)

set.seed(777)
data_split <- initial_split(new_df, prop = 0.8, strata = EVENT_TYPE)
train_data <- training(data_split)
test_data  <- testing(data_split)
 
cat("\nTrain size:", nrow(train_data), "| Test size:", nrow(test_data), "\n")
 
cat("\nTrain class distribution:\n")
train_data |> count(EVENT_TYPE, sort = TRUE) |> print()
```

# Model training -- Random Forest Training & Tuning

```{r, include = FALSE}
library(ranger)
```

```{r}
# set the recipe
rf_recipe <- recipe(EVENT_TYPE ~ ., data = train_data) |>
  step_zv(all_predictors())
```

```{r}

# setting model specifications (the recipe proportions)
rf_spec <- rand_forest(
  trees = 300,
  mtry  = tune(),
  min_n = tune()
) |>
  set_engine("ranger", max.depth = 15, importance = "impurity") |>
  set_mode("classification")
 
```

```{r}

# setting the workflow (how to make the recipe)
rf_workflow <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)

```

```{r}
# Cross-Validation Folds

set.seed(777)

# Stratified 5-fold cross-validation ensures each fold has a representative distribution of the EVENT_TYPE classes, which is crucial given the class imbalance.

cv_folds <- vfold_cv(train_data, v = 5, strata = EVENT_TYPE)

```

```{r, cache = TRUE}
# tuning

n_features <- ncol(train_data) - 1 # subtract 1 for the target variable
sqrt_features <- floor(sqrt(n_features))

# Grid search .
rf_grid <- grid_regular(
  mtry(range = c(sqrt_features, n_features)),
  min_n(range = c(5, 15)),
  levels = 4
)

# Macro f1 is the primary metric for evaluation due to the class imbalance.
tune_results <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(f_meas),
  control   = control_grid(verbose = FALSE)
)
```

```{r}
cat("\nTop 5 configurations by macro-F1:\n")
top_5_params <- tune_results |>
  show_best(metric = "f_meas", n = 5)

print(top_5_params)

best_params <- select_best(tune_results, metric = "f_meas")
cat("\nBest hyperparameters:\n")
print(best_params)

# use 7 and 6 instead
# manual_best_params <- tibble(
#   mtry = 11,
#   min_n = 6,
# )
```

```{r}
# fit worlflow with the best hyperparameters on the full training set
final_workflow <- rf_workflow |>
  finalize_workflow(best_params)
```

```{r, cache=TRUE}
cat("\nFitting final model on full training set...\n")
final_model <- final_workflow |>
  fit(data = train_data)
```



# Evaluation
```{r, include = FALSE}
library(vip)
```

```{r}
# add the predictions to the train and test for evaluation and reporting 

train_preds <- train_data |>
  bind_cols(predict(final_model, train_data))

# mutate event type to factor data type
train_preds <- train_preds |>
  mutate(EVENT_TYPE = factor(EVENT_TYPE))

test_preds <- test_data |>
  bind_cols(predict(final_model, test_data))

# mutate event type to factor data type
test_preds <- test_preds |>
  mutate(EVENT_TYPE = factor(EVENT_TYPE))
```


```{r}
# Accuracy check

# training set
train_accuracy <- mean(train_preds$.pred_class == train_preds$EVENT_TYPE)

bal_acc_train <- train_preds |>
  bal_accuracy(truth = EVENT_TYPE, estimate = .pred_class) |>
  pull(.estimate)

cat("Training Accuracy:  ", round(train_accuracy, 3), "\n")
cat("Balanced Training Accuracy: ", round(bal_acc_train, 3), "\n")


# test set
test_accuracy <- mean(test_preds$.pred_class == test_preds$EVENT_TYPE)

bal_acc_test <- test_preds |>
  bal_accuracy(truth = EVENT_TYPE, estimate = .pred_class) |>
  pull(.estimate)

cat("Test Accuracy:  ", round(test_accuracy, 3), "\n")
cat("Balanced Test Accuracy: ", round(bal_acc_test, 3), "\n")
```

```{r}
conf_mat_result <- test_preds |>
  conf_mat(truth = EVENT_TYPE, estimate = .pred_class)
print(conf_mat_result)
```


```{r}
# Per Class Prediction metrics
 
# Recall: the ability to find all actual positive cases (TP / TP + FN)
recall_tbl <- test_preds |>
  group_by(EVENT_TYPE) |>
  summarise(
    n       = n(),
    correct = sum(EVENT_TYPE == .pred_class),
    recall  = correct / n,
    .groups = "drop"
  )
 
# Precision: the accuracy of positive predictions (TP / TP + FP)
precision_tbl <- test_preds |>
  group_by(.pred_class) |>
  summarise(
    predicted_n = n(),
    true_pos    = sum(EVENT_TYPE == .pred_class),
    precision   = true_pos / predicted_n,
    .groups     = "drop"
  ) |>
  rename(EVENT_TYPE = .pred_class)
 
# Combine into a single report
class_report <- recall_tbl |>
  left_join(precision_tbl |> select(EVENT_TYPE, precision), by = "EVENT_TYPE") |>
  mutate(f1 = 2 * (precision * recall) / (precision + recall)) |>
  select(EVENT_TYPE, n, precision, recall, f1) |>
  arrange(desc(f1))
 
cat("\nPer-Class Metrics:\n")
print(class_report)
 
macro_f1    <- mean(class_report$f1, na.rm = TRUE)
weighted_f1 <- weighted.mean(class_report$f1, class_report$n, na.rm = TRUE)

cat("\nMacro-average F1:   ", round(macro_f1, 3), "\n")
cat("Weighted-average F1:", round(weighted_f1, 3), "\n")
```

```{r}
# Feature Importance
 
final_fitted <- final_model |> extract_fit_parsnip()
 
importance_tbl <- final_fitted |>
  vip::vi() |>
  arrange(desc(Importance))
 
cat("Feature Importances (Gini impurity):\n")
print(importance_tbl, n = 20)
```

```{r}
# Feature Importance Plot
 
importance_tbl |>
  mutate(Variable = fct_reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col(fill = "steelblue") +
  labs(
    title    = "Random Forest — Feature Importance",
    subtitle = "Storm Event Type Classification (Gini Impurity)",
    x        = "Importance",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)

```

```{r}
# Confusion Matrix Heatmap
 
cm_tbl <- conf_mat_result$table |>
  as_tibble() |>
  rename(Predicted = Prediction, Actual = Truth, Count = n)
 
cm_tbl |>
  ggplot(aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Count), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix — Random Forest",
    x     = "Predicted Class",
    y     = "Actual Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

```

