---
title: "Claasifier for Storm Event Type"
output: html_document
---

# Jason Ryan May

**Research Question:**

* Given where a storm occurs, when it occurs, what the local climate and demographics look like, and the storm's physical characteristics; Can we predict what type of severe weather event it is?


# Data Cleaning and Preprocessing

```{r, include=FALSE}
library(tidyverse)
library(tidymodels)
```

```{r}
# ── 1. Load Raw Data ────────────────────────────────────────────────────────
raw_df <- read_csv("mergedData_2023.csv", show_col_types = FALSE)
raw_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```

* It should be noted there is an imbalance of observations in each class. Additionally, there are only 257 lighting events, which is a very small number of observations to train a model on.

```{r}
# ── 2. Merge Lightning into Thunderstorm Wind ──────────────────────────────
#
# Lightning has too few observations (~256) and is nearly indistinguishable
# from Thunderstorm Wind using these features. Merging improves model
# performance on all other classes without meaningful information loss.
 
# new_df <- raw_df |>
#   mutate(EVENT_TYPE = if_else(EVENT_TYPE == "Lightning",
#                               "Thunderstorm Wind",
#                               EVENT_TYPE))
#  
# cat("\nClass distribution after merging Lightning into Thunderstorm Wind:\n")
# new_df |> count(EVENT_TYPE, sort = TRUE) |> print()
```


```{r}
# ── 3. Binary Indicators for Zero-Inflated Columns ─────────────────────────
#
# DAMAGE_PROPERTY (71.5% zeros), STORM_AREA_SQMILES (79.9% zeros), and
# DURATION_MINUTES (57.8% zeros) are heavily zero-inflated. Adding binary
# flags lets the model split on "zero vs nonzero" independently from
# the magnitude.
 
new_df <- raw_df |>
  mutate(
    DAMAGE_PROPERTY_NONZERO    = as.integer(DAMAGE_PROPERTY > 0),
    STORM_AREA_SQMILES_NONZERO = as.integer(STORM_AREA_SQMILES > 0),
    DURATION_MINUTES_NONZERO   = as.integer(DURATION_MINUTES > 0)
  )

# drop the original columns since the magnitude is not meaningful when zero-inflated
new_df <- new_df |>
  select(-DAMAGE_PROPERTY, -STORM_AREA_SQMILES, -DURATION_MINUTES)

# drop state, month_name, FIPS and cz_name since they are redundant with the lat and lon and month number
new_df <- new_df |>
  select(-STATE, -MONTH_NAME, -CZ_NAME, -FIPS)


# hot swap all categorical variables to numeric factors
new_df <- new_df |>
  mutate(
    EVENT_TYPE = factor(EVENT_TYPE),
    COASTAL_TYPE_SHORELINE = as.numeric(as.factor(COASTAL_TYPE_SHORELINE)),
    COASTAL_TYPE_WATERSHED = as.numeric(as.factor(COASTAL_TYPE_WATERSHED)),
    MODAL_YEAR_BUILT_BIN   = as.numeric(as.factor(MODAL_YEAR_BUILT_BIN))
    )

glimpse(new_df)
```

```{r}
# ── 5. Stratified Train / Test Split ───────────────────────────────────────

set.seed(42)
data_split <- initial_split(new_df, prop = 0.8, strata = EVENT_TYPE)
train_data <- training(data_split)
test_data  <- testing(data_split)
 
cat("\nTrain size:", nrow(train_data), "| Test size:", nrow(test_data), "\n")
 
cat("\nTrain class distribution:\n")
train_data |> count(EVENT_TYPE, sort = TRUE) |> print()
```

# Model training -- Random Forest Training & Tuning

```{r, include= = FALSE}
library(ranger)
```

```{r}
# ── 1. Preprocessing Recipe ────────────────────────────────────────────────
#
# All features are already numeric from the data prep step. The recipe
# only needs to remove any zero-variance predictors that might arise.

# set the recipe
rf_recipe <- recipe(EVENT_TYPE ~ ., data = train_data) |>
  step_zv(all_predictors())
```

```{r}
# ── 2. Model Specification ─────────────────────────────────────────────────
#
# Random Forest via ranger with two tunable hyperparameters:
#   - mtry:  number of features randomly sampled at each split
#   - min_n: minimum observations required in a terminal node
# trees is fixed at 300 (sufficient for convergence without excessive runtime).

# importance = "impurity" enables Gini-based feature importance extraction.

# use ranger for random forest implementation
rf_spec <- rand_forest(
  trees = 300,
  mtry  = tune(),
  min_n = tune()
) |>
  set_engine("ranger", importance = "impurity") |>
  set_mode("classification")
 
```

```{r}
# ── 4. Workflow ─────────────────────────────────────────────────────────────

# A workflow bundles the recipe and model specification together, ensuring that
# all preprocessing steps are correctly applied during tuning and final fitting.
rf_workflow <- workflow() |>
  add_recipe(rf_recipe) |>
  add_model(rf_spec)

```

```{r}
# ── 5. Cross-Validation Folds ──────────────────────────────────────────────

set.seed(42)

# Stratified 5-fold cross-validation ensures each fold has a representative distribution of the EVENT_TYPE classes, which is crucial given the class imbalance.
cv_folds <- vfold_cv(train_data, v = 5, strata = EVENT_TYPE)

```

```{r, cache = TRUE}
# ── 6. Hyperparameter Tuning ───────────────────────────────────────────────
#
# Grid search over mtry (4–10) and min_n (2–15) with 4 levels each,
# producing 16 configurations evaluated across 5 folds (80 fits total).
 

# A small grid is chosen to balance thoroughness with computational efficiency, especially given the size of the training set and the number of trees.
rf_grid <- grid_regular(
  mtry(range = c(4, 10)),
  min_n(range = c(2, 15)),
  levels = 4
)
 
cat("\nTuning", nrow(rf_grid), "configurations x 5 folds...\n")

# tune the model using the workflow, cross-validation folds, and tuning grid. Balanced accuracy is the primary metric for evaluation due to the class imbalance.
tune_results <- tune_grid(
  rf_workflow,
  resamples = cv_folds,
  grid      = rf_grid,
  metrics   = metric_set(bal_accuracy),
  control   = control_grid(verbose = FALSE)
)
 
cat("\nTop 5 configurations by balance accuracy:\n")
tune_results |>
  show_best(metric = "bal_accuracy", n = 5) |>
  print()
```


```{r}
# ── 7. Finalize & Fit Best Model ──────────────────────────────────────────
 
best_params <- select_best(tune_results, metric = "bal_accuracy")
cat("\nBest hyperparameters:\n")
print(best_params)
 
final_workflow <- rf_workflow |>
  finalize_workflow(best_params)
 
cat("\nFitting final model on full training set...\n")
final_model <- final_workflow |>
  fit(data = train_data)
 
cat("Done. Model fitted with", best_params$mtry, "mtry and",
    best_params$min_n, "min_n\n")
```


# Evaluation
```{r, include = FALSE}
library(vip)
```

```{r}
# ── 1. Predictions ─────────────────────────────────────────────────────────
test_preds <- test_data |>
  bind_cols(predict(final_model, test_data))
```


```{r}
# ── 2. Overall Accuracy ────────────────────────────────────────────────────
#
# Regular accuracy is dominated by the majority class (Thunderstorm Wind).
# Balanced accuracy averages recall across all classes equally, giving a
# fairer picture of performance on minority classes like Heavy Rain and
# Tornado.
 
test_accuracy <- mean(test_preds$.pred_class == test_preds$EVENT_TYPE)
 
bal_acc <- test_preds |>
  bal_accuracy(truth = EVENT_TYPE, estimate = .pred_class) |>
  pull(.estimate)

cat("Overall Accuracy:  ", round(test_accuracy, 3), "\n")
cat("Balanced Accuracy: ", round(bal_acc, 3), "\n")
```

```{r}
conf_mat_result <- test_preds |>
  conf_mat(truth = EVENT_TYPE, estimate = .pred_class)
print(conf_mat_result)
```


```{r}
# ── 3. Per-Class Precision, Recall, F1 ─────────────────────────────────────
 
# Recall: of all actual X, how many did we correctly predict as X?
recall_tbl <- test_preds |>
  group_by(EVENT_TYPE) |>
  summarise(
    n       = n(),
    correct = sum(EVENT_TYPE == .pred_class),
    recall  = correct / n,
    .groups = "drop"
  )
 
# Precision: of all predicted X, how many were actually X?
precision_tbl <- test_preds |>
  group_by(.pred_class) |>
  summarise(
    predicted_n = n(),
    true_pos    = sum(EVENT_TYPE == .pred_class),
    precision   = true_pos / predicted_n,
    .groups     = "drop"
  ) |>
  rename(EVENT_TYPE = .pred_class)
 
# Combine into a single report
class_report <- recall_tbl |>
  left_join(precision_tbl |> select(EVENT_TYPE, precision), by = "EVENT_TYPE") |>
  mutate(f1 = 2 * (precision * recall) / (precision + recall)) |>
  select(EVENT_TYPE, n, precision, recall, f1) |>
  arrange(desc(f1))
 
cat("\nPer-Class Metrics:\n")
print(class_report)
 
macro_f1    <- mean(class_report$f1, na.rm = TRUE)
weighted_f1 <- weighted.mean(class_report$f1, class_report$n, na.rm = TRUE)
 
cat("\nMacro-average F1:   ", round(macro_f1, 3), "\n")
cat("Weighted-average F1:", round(weighted_f1, 3), "\n")
```

```{r}
# ── 4. Feature Importance ──────────────────────────────────────────────────
 
final_fitted <- final_model |> extract_fit_parsnip()
 
importance_tbl <- final_fitted |>
  vip::vi() |>
  arrange(desc(Importance))
 
cat("Feature Importances (Gini impurity):\n")
print(importance_tbl, n = 20)
```

```{r}
# ── 5. Feature Importance Plot ─────────────────────────────────────────────
 
importance_tbl |>
  mutate(Variable = fct_reorder(Variable, Importance)) |>
  ggplot(aes(x = Importance, y = Variable)) +
  geom_col(fill = "steelblue") +
  labs(
    title    = "Random Forest — Feature Importance",
    subtitle = "Storm Event Type Classification (Gini Impurity)",
    x        = "Importance",
    y        = NULL
  ) +
  theme_minimal(base_size = 13)

```

```{r}
# ── 6. Confusion Matrix Heatmap ───────────────────────────────────────────
 
cm_tbl <- conf_mat_result$table |>
  as_tibble() |>
  rename(Predicted = Prediction, Actual = Truth, Count = n)
 
cm_tbl |>
  ggplot(aes(x = Predicted, y = Actual, fill = Count)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = Count), size = 3.5) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(
    title = "Confusion Matrix — Random Forest",
    x     = "Predicted Class",
    y     = "Actual Class"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))

```

